### 2019-10-03 manuscript.tex
develop the latex pole table

\section*{INTRODUCTION}

@@

\begin{figure}
\centering
\includegraphics[width=4in]{Figures/Tectonic_history.pdf}
\caption{\small{\textbf{Summary of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}}
\label{fig:tectonic_history}
\end{figure}

@@

In this chapter, I will focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications. The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results. While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2013a}

Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean cratons. There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation. These data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly. Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia. A lingering question raise in \citet{Hoffman198a} that still lingers is whether the Archean provicnes each had independent drit histories with significant speraation ro whether they have shared historieds before experiencing fragmantation and remamalgamation.

For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior province relative to the eastern Superior province across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}. This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny. \cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga.  I follow this interpretation and group the poles into Superior (West) and Superior (East). These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame. Uncertainty remains with respect to whether the ca. 1.88 Ga Molson dikes pole pre-dates or post-dates this rotation and thus for the time-being should be considered solely in the western Superior province reference grame.

Following Laurentia's amalgamation, poles from each part of the Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls201Xa} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}}). Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the following rotations:

\begin{itemize}
\item Laurentia-Greenland.  1202 1800.0   67.5 -118.5  -13.8  1000 !Rotation of Greenland to Laurentia
\item Laurentia-Scotland 1022 1800.0   73.0   96.5  -22.0  1000 !Rotate (proto-)Scotland to Laurentia-Greenland fit according to Darabi2004
\item Laurentia-Svalbard 3301 1275.0  -81.0  125.0   68.0  1000 !Rotate Svalbard to Laurentia in fit that works well with East Greenland basin (Maloof et al., 2006 rotation)
\end{itemize}

>>>>

\section*{Introduction}

@@

\begin{figure}
\centering
\includegraphics[width=4in]{Figures/Tectonic_history.pdf}
\caption{\small{\textbf{Simplified summary of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. Note that the Penokean orogeny overlaps with the Trans-Hudson orogeny and is not shown.}}
\label{fig:tectonic_history}
\end{figure}

@@

In this chapter, I focus on the compilation of paleomagnetic poles developed through the Nordic Paleomagnetism Workshops with some additions and modifications. The Nordic Paleomagnetism Workshops have taken the approach of using expert panels to assess paleomagnetic poles and assign them grades meant to convey the confidence that the community has in these results. While many factors associated with paleomagnetic poles can be assessed quantitatively such as the Fisher statistics and the precision of geochronological constraints, other aspects such as the degree to which available field tests constrain the magnetization to be primary require expert assessment. The categorizations used by the expert panel are `A' and `B' with the last panel meeting occurring in Fall 2017 in Leirubakki, Iceland. An `A' rating refers to poles that are judged to be of such high-quality that they provide essential constraints that should be satisfied in paleogeographic reconstructions. A `B' rating is associated with poles that are judged to likely provide a high-quality constraint, but have some deficiency such as remaining ambiguity in the demonstration of primary remanence or the quality/precision of available geochronologic constraints. Additional poles that were not given an `A' or `B' classification at the Nordic Workshops are referred to as not-rated (`NR'). These additional poles are taken from the Paleomagia database \citep{Veikkolainen2013a}. Many of these poles are quite valuable for reconstruction and should not be dismissed from being considered in paleogeography reconstructions. For example, there are rich data associated with intrusive lithologies of the Grenville Province that are the available paleomagnetic constraints for Laurentia at the Mesoproterozoic-Neoproterozoic boundary. However, the ages of the remanence associated with these poles is complicated by the reality that the magnetization was acquired during exhumation and such cooling ages are more difficult to robustly constrain than for eruptive units or shallow-level intrusions. As a result, the vast majority of Grenville Province poles are not given an `A' or `B' rating. However, while any one of these Grenville poles could be interpreted to suffer from temporal uncertainty, the overall preponderence of poles in a similar location at the time suggests that they need to be taken seriously within any paleogeographic reconstruction of Laurentia (although note an alternative view of an allochthnous origin discussed below).

Prior to the termination of the Trans-Hudson orogeny (before 1.8 Ga), I consider paleomagnetic poles with respect to the individual Archean provinces. There are poles in the compilation for the Slave, Wyoming, Rae, Superior and Nain provinces prior to Laurentia amalgamation. A lingering question raised in \citet{Hoffman198a} that still lingers is whether the Archean provinces each had independent drift histories with significant seperation or whether they have shared historieds before experiencing fragmantation and remamalgamation. Overall, these data provide an opportunity to re-evaluate the paleomagnetic evidence for relative motions between Archean provinces prior to Laurentia assembly. Following the termination of this orogeny (after 1.8 Ga), I consider poles from all the respective provinces and terranes to reflect the position of all of Laurentia.

For the Superior province, an additional complexity is that paleomagnetic poles from Siderian to Rhyacian Period (2.50 to 2.05 Ga) dike swarms, as well as deflection of dike trends, support an interpretation that there was substantial Paleoproterozoic rotation of the western Superior province relative to the eastern Superior province across the Kapuskasing Structural Zone \citep{Bates1991a, Evans2010a}. This interpretation is consistent with the hypothesis of \citet{Hoffman1988a} that Kapuskasing Structural Zone represents major intracratonic uplift related to the Trans-Hudson orogeny. \cite{Evans2010a} propose an Euler rotation of (51\textdegree N, 85\textdegree W, -14\textdegree CCW) to reconstruct western Superior relative to eastern Superior and interpret that the rotation occured in the time interval of 2.07 to 1.87 Ga.  I follow this interpretation and group the poles into Superior (West) and Superior (East). These poles are shown in Table 1 both in their initial reference frame and rotated into the other Superior reference frame. Uncertainty remains with respect to whether the ca. 1.88 Ga Molson dikes pole pre-dates or post-dates this rotation and thus for the time-being should be considered solely in the western Superior province reference frame.

Following Laurentia's amalgamation, poles from each part of the Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015a} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochothon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction and I apply the following rotations:

\begin{itemize}
\item Laurentia-Greenland.  1202 1800.0   67.5 -118.5  -13.8  1000 !Rotation of Greenland to Laurentia
\item Laurentia-Scotland 1022 1800.0   73.0   96.5  -22.0  1000 !Rotate (proto-)Scotland to Laurentia-Greenland fit according to Darabi and Piper (2004)
\item Laurentia-Svalbard 3301 1275.0  -81.0  125.0   68.0  1000 !Rotate Svalbard to Laurentia in fit that works well with East Greenland basin (Maloof et al., 2006 rotation)
\end{itemize}

{\scriptsize
\begin{landscape}
\input{pole_table_editted.tex}
\end{landscape}
}

These paleomagnetic poles can be used to develop paleogeographic reconstructions for Laurentia alone. Paleogeographic snapshots for the past position of Laurentia are shown in Figure \ref{fig:recons_time}. These reconstructions use the tectonic elements as

As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block). While they provide constraints in this regard, they do not provide constraints in and of themselves
